Effective multi-body SU($N$)-symmetric interactions of ultracold fermionic atoms on a 3-D lattice

### 2019-02-11 multi_body_interactions/papers/theory/reviews_NJP/referee_1_response.tex

\begin{document}

\section*{Referee 1}

We thank the referee for taking the time to carefully read our
manuscript, and for providing thoughtful and constructive feedback.
Below, we address the comments and concerns brought up by the referee.
We hope that the referee finds our responses and revisions
satisfactory, deeming our manuscript acceptable for publication in New
Journal of Physics.

Excerpts from the referee reports are written in \blue{blue}, excerpts
from our original manuscript are written in \red{red}, and excerpts
from the revised version of our manuscript is written in
\green{green}.  All page, line, equation, and reference numbers
generally refer to those in our original manuscript, unless explicitly
stated otherwise.

@@

  We also clarify the limitations of our theory in the text between
  Eqs.~(8) and (9), as discussed later in point \ref{pt:confusion},
  and have made our technical contributions more explicit in the
  description of our main results (in section II), replacing (page 7,
  line 48):

  \red{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} H_M, \tag{9}
    \end{align*}
    where $H_M$ is an $M$-body Hamiltonian, $I$ is the total nuclear
    spin of each atom (e.g.~$I=9/2$ for ${}^{87}$Sr), and the sum
    terminates at $2I+1$ because this is the largest number of atoms
    which may initially occupy a single lattice site.  Focusing on the
    low-lying orbital state excitations of this theory, we find that
    SU($N$) symmetry allows us to express multi-body Hamiltonians in
    the simple form}

  by:

  \green{Our low-energy effective theory exhibits SU($N$)-symmetric
    multi-body interactions, such that the effective interaction
    Hamiltonian can be written in the form
    \begin{align*}
      H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
      \tag{9}
    \end{align*}
    where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
    coupling constants $G_X$, and $I$ is the total nuclear spin of
    each atom (e.g.~$I=9/2$ for ${}^{87}$Sr).  The sum terminates at
    $2I+1$ because this is the largest number of atoms which may
    initially occupy a single lattice site.  We explicitly compute all
    of the $M$-body Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through
    order $p=3$, yielding effective two-, three-, and four-body
    interactions.  To (i) facilitate a comparison with the
    experimental measurements of many-body orbital excitation spectra
    performed in ref.~[34] and (ii) characterize the low-lying
    excitations in our effective theory, we additionally restrict the
    multi-body Hamiltonians $H_M$ to states with at most one orbital
    excitation per lattice site.  Under this restriction, we find that
    the SU($N$) symmetry of atomic collisions allows us to express all
    multi-body Hamiltonians in the simple form}

@@

  \red{Consider a collection of fermionic atoms with total nuclear
    spin $I$ and two atomic orbital states loaded into a
    translationally invariant 3-D optical lattice.  While an external
    trapping potential will generally break discrete translational
    symmetry of the lattice, any background inhomogeneity can be made
    negligible by spectroscopically addressing a sufficiently small
    region of the lattice [34].  Throughout this paper, we work
    strictly in the deep-lattice regime with negligible tunneling
    between lattice sites.  We also assume a so-called
    ``magic-wavelength'' lattice for which both atomic orbital states
    experience identical lattice potentials [40].}

  now reads:

  \green{The work in this paper is closely tied to the experimental
    work reported in ref.~[34]; we begin with a short summary of the
    relevant experimental procedures therein.  The experiment begins
    by preparing a degenerate gas of $10^4$-$10^5$ (fermionic)
    ${}^{87}$Sr atoms in a uniform mixture of their ten nuclear spin
    states and at $\sim0.1$ of their fermi temperature ($\sim10$
    nanokelvin) [8, 40].  This gas is loaded into a primitive cubic
    optical lattice at the ``magic wavelength'' for which both ground
    (${}^1S_0$) and first-excited (${}^3P_0$) electronic orbital
    states of the atoms experience the same lattice potential [41].
    Lattice depths along the principal axes of the lattice are roughly
    equal in magnitude, with a geometric mean that can be varied from
    30 to 80 $E_{\t{R}}$, where
    $E_{\t{R}}\approx3.5\times2\pi~\t{kHz}$ is the lattice photon
    recoil energy of the atoms (with the reduced Planck constant
    $\hbar=1$ throughout this paper).  These lattice depths are
    sufficiently large as to neglect tunneling on the time scales
    relevant to the experiment.  The temperature of the atoms is also
    low enough to neglect thermal occupation of motional states
    outside the ground-state manifold.}

  \green{Once loaded into an optical lattice, atoms are addressed by
    an external (``clock'') interrogation laser with an ultranarrow
    (26 mHz) linewidth, detuned by $\Delta$ from the single-atom
    ${}^1S_0-{}^3P_0$ transition frequency $\omega_0$.  After a fixed
    interrogation time, the experiment turns off the interrogation
    laser, removes all ground-state (${}^1S_0$) atoms from the
    lattice, and uses absorption imaging to count the remaining
    excited-state (${}^3P_0$) atoms.  Non-interacting atoms in
    singly-occupied lattice sites feature the typical single particle
    lineshape peaked at $\Delta=0$.  The lineshapes of
    multiply-occupied lattice sites, meanwhile, are shifted by
    inter-atomic interactions, which results in spectroscopic peaks
    (i.e.~local maxima in excited-state atom counts) away from
    $\Delta=0$.  A sweep across different detunings $\Delta$ (on the
    scale of inter-atomic interaction energies) thus constitutes a
    measurement of the many-body orbital excitation spectrum.  We note
    that this spectroscopic protocol addresses only singly-excited
    orbital states of lattice sites.  Doubly-excited states are
    off-resonant due to (i) the interaction-induced non-linearity
    ($\sim$kHz) of orbital excitation energies, and (ii) the
    ultranarrow linewidth ($\sim$mHz) of the interrogation laser.}

### 2019-02-11 multi_body_interactions/papers/theory/paper.tex
final changes to referee 1 response (NJP)

Once loaded into an optical lattice, atoms are addressed by an
external (``clock'') interrogation laser with an ultranarrow (26 mHz)
linewidth, detuned by $\Delta$ from the single-atom ${}^1S_0-{}^3P_0$
transition frequency $\omega_0$.  After a fixed interrogation time,
the experiment turns off the interrogation laser, removes all
ground-state (${}^1S_0$) atoms from the lattice, and uses absorption
imaging to count the remaining excited-state (${}^3P_0$) atoms.
Non-interacting atoms in singly-occupied lattice sites feature the
typical single particle lineshape peaked at $\Delta=0$.  The
lineshapes of multiply-occupied lattice sites, meanwhile, are shifted
by inter-atomic interactions, which results in spectroscopic peaks
(i.e.~local maxima in excited-state atom counts) away from $\Delta=0$.
A sweep across different detunings $\Delta$ (on the scale of
inter-atomic interaction energies) thus constitutes a measurement of
the many-body orbital excitation spectrum.  We note that this
spectroscopic protocol addresses only singly-excited orbital states of
lattice sites.  Doubly-excited states are manifestly off-resonant for
single-photon absorption processes as $\abs{\Delta}\ll\omega_0$, while
multi-photon absorption processes are always off-resonant due to
\begin{enumerate*}
\item the interaction-induced non-linearity ($\sim$kHz) of the orbital
  excitation energies, and
\item the ultranarrow linewidth ($\sim$mHz) of the interrogation
  laser.
\end{enumerate*}

@@

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
\end{align}
where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
coupling constants $G_X$, and $I$ is the total nuclear spin of each
atom (e.g.~$I=9/2$ for ${}^{87}$Sr).  The sum terminates at $2I+1$
because this is the largest number of atoms which may initially occupy
a single lattice site.  We explicitly compute all $M$-body
Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through order $p=3$, yielding
effective two-, three-, and four-body interactions.  To
\begin{enumerate*}
\item facilitate a comparison with the experimental measurements of
  many-body orbital excitation spectra performed in
  ref.~\cite{goban2018emergence} andr
\item characterize the low-lying excitations in our effective theory,
\end{enumerate*}
we also consider a restriction of the multi-body Hamiltonians $H_M$ to
states with at most one orbital excitation per lattice site.  Under
this restriction, we find that the SU($N$) symmetry of atomic
collisions allows us to express all multi-body Hamiltonians in the
simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  H_2^{(\mu_1,\mu_2)} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g},
  \label{eq:H_M_preview}
\end{align}
where $H_2^{(\mu_1,\mu_2)}$ is a two-body Hamiltonian addressing atoms
with nuclear spin $\mu_1,\mu_2$; and
$\n_{\mu s}=\c_{\mu s}^\dag\c_{\mu s}$ is a number operator for atoms
with nuclear state $\mu$ and orbital state $s$.  The sum in
\eqref{eq:H_M_preview} is performed over all choices of nuclear spins
$\mu_j$ with $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The key feature of the $M$-body
interactions in \eqref{eq:H_M_preview} is that they ultimately take
the same form as two-body interactions, but with the addition of $M-2$
spectator atoms.  This form is a direct consequence of the SU($N$)
symmetry of underlying two-body interactions.

@@

Using the machinery developed in ref.~\cite{bravyi2011schrieffer} for
performing a rotation between low-energy subspaces of a perturbed
(i.e.~interacting) and unperturbed (i.e.~non-interacting) Hamiltonian,
we derive an expansion for an effective interaction Hamiltonian
$H_{\t{int}}^{\t{eff}}$ in terms two-body interaction Hamiltonian
$H_{\t{int}}$ (see Appendix \ref{sec:eff_derivation}).  This expansion
takes the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{p\ge1} H_{\t{int}}^{(p)},
  \label{eq:H_int_eff}
\end{align}
where $H_{\t{int}}^{(p)}$ is order $p$ in $H_{\t{int}}$.  Letting
$\E_0\equiv \H_{\t{full}}\setminus\H_{\t{ground}}^{\t{single}}$ denote
the orthogonal complement of $\H_{\t{ground}}^{\t{single}}$
(i.e.~$\E_0$ is the space of all states with at least one atom in an
excited motional state), $\B_0\p{\E_0}$ denote an eigenbasis of $\E_0$
with respect to the single-particle Hamiltonian $H_0$, and $E_\alpha$
denote the motional energy (with respect to $H_0$) of a state
$\ket\alpha\in\E_0$ relative to the corresponding motional
ground-state energy, we define the operator
\begin{align}
  \I \equiv \sum_{\ket\alpha\in\B_0\p{\E_0}} \f{\op\alpha}{E_\alpha},
\end{align}
which sums over projections onto excited states with corresponding
energetic suppression factors.  The operator $\I$ together with the
projector $\P_0$ onto $\H_{\t{ground}}^{\t{single}}$ allows us
concisely write the first few terms in \eqref{eq:H_int_eff} as
\begin{align}
  H_{\t{int}}^{(1)} = \P_0 H_{\t{int}} \P_0,
  &&
  H_{\t{int}}^{(2)} = -\P_0 H_{\t{int}} \I H_{\t{int}} \P_0,
  \label{eq:H_int_1_2}
\end{align}
\begin{align}
  H_{\t{int}}^{(3)}
  = \P_0 H_{\t{int}} \I H_{\t{int}} \I H_{\t{int}} \P_0
  - \f12\sp{\P_0 H_{\t{int}} \P_0,
    \P_0 H_{\t{int}} \I^2 H_{\t{int}} \P_0}_+,
  \label{eq:H_int_3}
\end{align}
where $\sp{X,Y}_+\equiv XY+YX$.  Writing down a single-band Hubbard
model that simply neglects excited atomic motional states and uses
$H_{\t{int}}$ directly to describe the orbital spectrum of interacting
atoms is thus equivalent to truncating our expansion for
$H_{\t{int}}^{\t{eff}}$ at first order.  In addition to this first
order term, the expansion involves {\it effective corrections} to the
action of $H_{\t{int}}$ on the single-particle motional ground states
(i.e.~on $\H_{\t{ground}}^{\t{single}}$) in the form of higher-order
terms with {\it intermediate} or {\it virtual occupation} of excited
states, manifest in $\I$.

@@

The introduction of counter-terms is merely a formal decomposition of
the ``bare'' coupling constants $G_X^{\t{bare}}$ that are used in
\eqref{eq:H_2_expansion} as $G_X^{\t{bare}} = G_X + \tilde G_X$.  In
performing such a decomposition, we are free to choose the values of
$G_X$, which in turn fixes the values of
$\tilde G_X \equiv G_X^{\t{bare}} - G_X$.  For convenience, we can
choose the values of $G_X$ to be those of the net effective coupling
constants on the right-hand side of \eqref{eq:H_2_expansion}.
Representing the new coupling constants $G_X$ by regular vertices and
the counter-terms $\tilde G_X$ by a crossed dot (i.e.~$\otimes$), this
choice leads to the {\it renormalization condition}
\begin{align}
  \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v);
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G)}
  + \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex[crossed dot] (v) {};
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(\tilde G)}
  - \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v1);
          \vertex[above left = of v1] (f1);
          \vertex[below left = of v1] (f2);
          \vertex[right = of v1] (v2);
          \vertex[above right = of v2] (f3);
          \vertex[below right = of v2] (f4);
          \diagram* {
            (f1) -- (v1),
            (f2) -- (v1),
            (v2) -- (f3),
            (v2) -- (f4),
            (v1) --[scalar, half left] (v2),
            (v1) --[scalar, half right] (v2) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G^2)}
  + \cdots = \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v);
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G)}.
  \label{eq:counter_term}
\end{align}
This renormalization condition has the benefit of allowing us to
express effective two-body interactions simply in terms of net
effective two-body coupling constants, rather than in terms of long
sums at all order of the bare coupling constants.  By construction,
the counter-terms we have introduced exactly cancel all terms beyond
leading order in \eqref{eq:H_2_expansion}, which implies that the
effective two-body interaction Hamiltonian is simply
\begin{align}
  H_2 = \shrink{
    \begin{tikzpicture}
      \begin{feynman}
        \vertex (v);
        \vertex[above left = of v] (f1);
        \vertex[below left = of v] (f2);
        \vertex[above right = of v] (f3);
        \vertex[below right = of v] (f4);
        \diagram* {
          (f1) -- (v),
          (f2) -- (v),
          (v) -- (f3),
          (v) -- (f4), };
      \end{feynman}
    \end{tikzpicture}}
  = \f12 \alpha_2^{(1)} \sum_{\abs{\set{\mu,\nu}}=2} G^{qr}_{st}
  \c_{\mu s}^\dag \c_{\nu t}^\dag \c_{\nu r} \c_{\mu q},
  \label{eq:H_2}
\end{align}
where for consistency with existing literature
\cite{johnson2012effective} we define $\alpha_2^{(1)} \equiv K$ as the
overlap integral between two atoms occupying single-particle motional
ground states.

@@

In addition to spectral measurements of many-body interaction
energies, we consider the dynamics of multiply-occupied lattice sites
during spectroscopic interrogation.  While these dynamics do not
provide information about the nature or origin of effective multi-body
interactions, they provide tools and intuition for addressing the
low-lying orbital excitations which are readily accessible in an
experimental setting.  If we initialize all atoms in the $N$-body
ground state with an incoherent mixture of all nuclear spins, then we
prepare the mixed state $\rho_{N,0}=\P_{N,0}/\Tr\P_{N,0}$, where
\begin{align}
  \P_{NX} \equiv \sum_{\abs{\N}=N}\op{\N X}
  \label{eq:P_NX}
\end{align}
is a projector onto the space of the $N$-body orbital states
$\ket{\N X}$.  Interrogating the atoms for a time $t$ by a laser
resonant with the excitation energy $\Delta_{N,\pm}$ then gives us the
state
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}}
  \sum_{\abs{\N}=N} \exp\p{-it\Omega_{\N,\pm}S_{\N,\pm}^\x}
  \op{\N,0} \exp\p{it\Omega_{\N,\pm}S_{\N,\pm}^\x},
\end{align}
where the Rabi frequencies $\Omega_{\N,+},\Omega_{\N,-}$ and
pseudo-spin operators $S_{\N,\pm}^\x$ are respectively given in
\eqref{eq:O_N+}, \eqref{eq:O_N-}, and \eqref{eq:S_NX_x}.  Denoting the
eigenstates of $S_{\N,\pm}^\x$ by
\begin{align}
  \ket{\N,\S_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} + \ket{\N,\pm}},
  &&
  \ket{\N,\A_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} - \ket{\N,\pm}},
\end{align}
and defining the identity operator projected to the relevant subspace,
\begin{align}
  \1_{\N,\pm} \equiv \op{\N,0} + \op{\N,\pm}
  = \op{\N,\S_\pm} + \op{\N,\A_\pm},
\end{align}
we can write the state $\rho_N^{(\pm)}\p{t}$ and excited-state
projectors $\P_{N,\pm}$ in the form
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}} \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N,\pm} + e^{i2t\Omega_{\N,\pm}}\op{\N,\S_\pm}{\N,\A_\pm}
    + e^{-i2t\Omega_{\N,\pm}}\op{\N,\A_\pm}{\N,\S_\pm}},
  \label{eq:rho_N_pm_t}
\end{align}
and
\begin{align}
  \P_{N,\pm} = \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N,\pm}-\op{\N,\S_\pm}{\N,\A_\pm}-\op{\N,\A_\pm}{\N,\S_\pm}},
\end{align}
from which it follows that the net excited-state population at time
$t$ is
\begin{align}
  \bk{\P_{N,\pm}\p{t}}
  \equiv \Tr\sp{\rho_N^{(\pm)}\p{t}\P_{N,\pm}}
  = \f12 - \f12 \bk{\cos\p{2t\Omega_{\N,\pm}}}_{\abs{\N}=N},
  \label{eq:P_N_pm_t}
\end{align}
where $\bk{X}_{\abs{\N}=N}\equiv\sum_{\abs{\N}=N}X/\Tr\P_{N,0}$ is an
average of $X$ over all choices of $N$ distinct nuclear spins.

>>>>

Once loaded into an optical lattice, atoms are addressed by an
external (``clock'') interrogation laser with an ultranarrow (26 mHz)
linewidth, detuned by $\Delta$ from the single-atom ${}^1S_0-{}^3P_0$
transition frequency $\omega_0$.  After a fixed interrogation time,
the experiment turns off the interrogation laser, removes all
ground-state (${}^1S_0$) atoms from the lattice, and uses absorption
imaging to count the remaining excited-state (${}^3P_0$) atoms.
Non-interacting atoms in singly-occupied lattice sites feature the
typical single particle lineshape peaked at $\Delta=0$.  The
lineshapes of multiply-occupied lattice sites, meanwhile, are shifted
by inter-atomic interactions, which results in spectroscopic peaks
(i.e.~local maxima in excited-state atom counts) away from $\Delta=0$.
A sweep across different detunings $\Delta$ (on the scale of
inter-atomic interaction energies) thus constitutes a measurement of
the many-body orbital excitation spectrum.  We note that this
spectroscopic protocol addresses only singly-excited orbital states of
lattice sites.  Doubly-excited states are off resonant due to
\begin{enumerate*}
\item the interaction-induced non-linearity ($\sim$kHz) of the orbital
  excitation energies, and
\item the ultranarrow linewidth ($\sim$mHz) of the interrogation
  laser.
\end{enumerate*}

@@

Our low-energy effective theory exhibits SU($N$)-symmetric multi-body
interactions, such that the effective interaction Hamiltonian can be
written in the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{M=2}^{2I+1} \sum_{p\ge1} H_M^{(p)},
\end{align}
where $H_M^{(p)}$ is an $M$-body Hamiltonian of order $p$ in the
coupling constants $G_X$, and $I$ is the total nuclear spin of each
atom (e.g.~$I=9/2$ for ${}^{87}$Sr).  The sum terminates at $2I+1$
because this is the largest number of atoms which may initially occupy
a single lattice site.  We explicitly compute all $M$-body
Hamiltonians $H_M\equiv\sum_p H_M^{(p)}$ through order $p=3$, yielding
effective two-, three-, and four-body interactions.  To
\begin{enumerate*}
\item facilitate a comparison with the experimental measurements of
  many-body orbital excitation spectra performed in
  ref.~\cite{goban2018emergence} and
\item characterize the low-lying excitations in our effective theory,
\end{enumerate*}
we additionally restrict the multi-body Hamiltonians $H_M$ to states
with at most one orbital excitation per lattice site.  Under this
restriction, we find that the SU($N$) symmetry of atomic collisions
allows us to express all multi-body Hamiltonians in the simple form
\begin{align}
  H_M = \sum_{\abs{\set{\mu_j}}=M}
  H_2^{(\mu_1,\mu_2)} \prod_{\alpha=3}^M \n_{\mu_\alpha,\g},
  \label{eq:H_M_preview}
\end{align}
where $H_2^{(\mu_1,\mu_2)}$ is a two-body Hamiltonian addressing atoms
with nuclear spin $\mu_1,\mu_2$; and
$\n_{\mu s}=\c_{\mu s}^\dag\c_{\mu s}$ is a number operator for atoms
with nuclear state $\mu$ and orbital state $s$.  The sum in
\eqref{eq:H_M_preview} is performed over all choices of nuclear spins
$\mu_j$ with $j=1,2,\cdots,M$ for which all $\mu_j$ are distinct, or
equivalently all choices of $\mu_j$ for which the set $\set{\mu_j}$
contains $M$ elements, for a total of ${2I+1\choose M}\times\p{M!}$
nuclear spin combinations.  The key feature of the $M$-body
interactions in \eqref{eq:H_M_preview} is that they ultimately take
the same form as two-body interactions, but with the addition of $M-2$
spectator atoms.  This form is a direct consequence of the SU($N$)
symmetry of underlying two-body interactions.

@@

Using the machinery developed in ref.~\cite{bravyi2011schrieffer} for
performing a rotation between low-energy subspaces of a perturbed
(i.e.~interacting) and unperturbed (i.e.~non-interacting) Hamiltonian,
we derive an expansion for an effective interaction Hamiltonian
$H_{\t{int}}^{\t{eff}}$ in terms two-body interaction Hamiltonian
$H_{\t{int}}$ (see Appendix \ref{sec:eff_derivation}).  This expansion
takes the form
\begin{align}
  H_{\t{int}}^{\t{eff}} = \sum_{p\ge1} H_{\t{int}}^{(p)},
  \label{eq:H_int_eff}
\end{align}
where $H_{\t{int}}^{(p)}$ is order $p$ in $H_{\t{int}}$.  Letting
$\E_0\equiv \H_{\t{full}}\setminus\H_{\t{ground}}^{\t{single}}$ denote
the orthogonal complement of $\H_{\t{ground}}^{\t{single}}$
(i.e.~$\E_0$ is the space of all states with at least one atom in an
excited motional state), $\B_0\p{\E_0}$ denote an eigenbasis of $\E_0$
with respect to the single-particle Hamiltonian $H_0$, and $E_\alpha$
denote the motional energy (with respect to $H_0$) of a state
$\ket\alpha\in\E_0$ relative to the corresponding motional
ground-state energy, we define the operator
\begin{align}
  \I \equiv \sum_{\ket\alpha\in\B_0\p{\E_0}} \f{\op\alpha}{E_\alpha},
\end{align}
which sums over projections onto excited states with corresponding
energetic suppression factors.  The operator $\I$ together with the
projector $\P_0$ onto $\H_{\t{ground}}^{\t{single}}$ allows us
concisely write the first few terms in \eqref{eq:H_int_eff} as
\begin{align}
  H_{\t{int}}^{(1)} = \P_0 H_{\t{int}} \P_0,
  &&
  H_{\t{int}}^{(2)} = -\P_0 H_{\t{int}} \I H_{\t{int}} \P_0,
  \label{eq:H_int_1_2}
\end{align}
\begin{align}
  H_{\t{int}}^{(3)}
  = \P_0 H_{\t{int}} \I H_{\t{int}} \I H_{\t{int}} \P_0
  - \f12\sp{\P_0 H_{\t{int}} \P_0,
    \P_0 H_{\t{int}} \I^2 H_{\t{int}} \P_0}_+,
  \label{eq:H_int_3}
\end{align}
where $\sp{X,Y}_+\equiv XY+YX$.  Writing down a single-band Hubbard
model that simply neglects excited atomic motional states and uses
$H_{\t{int}}$ directly to describe the orbital spectrum of interacting
atoms is thus equivalent to truncating our expansion for
$H_{\t{int}}^{\t{eff}}$ at first order.  In addition to this first
order term, the expansion involves {\it effective corrections} to the
action of $H_{\t{int}}$ on the non-interacting motional ground states
(i.e.~on $\H_{\t{ground}}^{\t{single}}$) in the form of higher-order
terms with {\it intermediate} or {\it virtual occupation} of excited
states, manifest in $\I$.

@@

The introduction of counter-terms is merely a formal decomposition of
the ``bare'' coupling constants $G_X^{\t{bare}}$ that are used in
\eqref{eq:H_2_expansion} as $G_X^{\t{bare}} = G_X + \tilde G_X$.  In
performing such a decomposition, we are free to choose the values of
$G_X$, which in turn fixes the values of
$\tilde G_X \equiv G_X^{\t{bare}} - G_X$.  For convenience, we can
choose the values of $G_X$ to be those of the net effective coupling
constants on the right-hand side of \eqref{eq:H_2_expansion}.
Representing the new coupling constants $G_X$ by regular vertices and
the counter-terms $\tilde G_X$ by a crossed dot (i.e.~$\otimes$), this
choice leads to the {\it renormalization condition}
\begin{align}
  \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v);
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G)}
  + \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex[crossed dot] (v) {};
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(\tilde G)}
  - \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v1);
          \vertex[above left = of v1] (f1);
          \vertex[below left = of v1] (f2);
          \vertex[right = of v1] (v2);
          \vertex[above right = of v2] (f3);
          \vertex[below right = of v2] (f4);
          \diagram* {
            (f1) -- (v1),
            (f2) -- (v1),
            (v2) -- (f3),
            (v2) -- (f4),
            (v1) --[scalar, half left] (v2),
            (v1) --[scalar, half right] (v2) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G^2)}
  + \cdots = \substack{
    \shrink{
      \begin{tikzpicture}
        \begin{feynman}
          \vertex (v);
          \vertex[above left = of v] (f1);
          \vertex[below left = of v] (f2);
          \vertex[above right = of v] (f3);
          \vertex[below right = of v] (f4);
          \diagram* {
            (f1) -- (v),
            (f2) -- (v),
            (v) -- (f3),
            (v) -- (f4) };
        \end{feynman}
      \end{tikzpicture}}
    \\ \O(G)}.
  \label{eq:counter_term}
\end{align}
This renormalization condition has the benefit of allowing us to
express effective two-body interactions simply in terms of net
effective two-body coupling constants, rather than in terms of long
sums at all order of the bare coupling constants.  By construction,
the counter-terms we have introduced exactly cancel all terms beyond
leading order in \eqref{eq:H_2_expansion}, which implies that the
effective two-body interaction Hamiltonian is simply
\begin{align}
  H_2 = \shrink{
    \begin{tikzpicture}
      \begin{feynman}
        \vertex (v);
        \vertex[above left = of v] (f1);
        \vertex[below left = of v] (f2);
        \vertex[above right = of v] (f3);
        \vertex[below right = of v] (f4);
        \diagram* {
          (f1) -- (v),
          (f2) -- (v),
          (v) -- (f3),
          (v) -- (f4), };
      \end{feynman}
    \end{tikzpicture}}
  = \f12 \alpha_2^{(1)} \sum_{\abs{\set{\mu,\nu}}=2} G^{qr}_{st}
  \c_{\mu s}^\dag \c_{\nu t}^\dag \c_{\nu r} \c_{\mu q},
  \label{eq:H_2}
\end{align}
where for consistency with existing literature
\cite{johnson2012effective} we define $\alpha_2^{(1)} \equiv K$ as the
overlap integral between two atoms occupying non-interacting motional
ground states.

@@

In addition to spectral measurements of many-body interaction
energies, we consider the dynamics of multiply-occupied lattice sites
during spectroscopic interrogation.  While these dynamics do not
provide information about the nature or origin of effective multi-body
interactions, they provide tools and intuition for addressing the
low-lying orbital excitations which are readily accessible in an
experimental setting.  If we initialize all atoms in the $N$-body
ground state with an incoherent mixture of all nuclear spins, then we
prepare the mixed state $\rho_{N,0}=\P_{N,0}/\Tr\P_{N,0}$, where
\begin{align}
  \P_{NX} \equiv \sum_{\abs{\N}=N}\op{\N X}
  \label{eq:P_NX}
\end{align}
is a projector onto the space of the $N$-body orbital states
$\ket{\N X}$.  Interrogating the atoms for a time $t$ by a laser
resonant with the excitation energy $\Delta_{N\pm}$ then gives us the
state
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}}
  \sum_{\abs{\N}=N} \exp\p{-it\Omega_{\N\pm}S_{\N\pm}^\x}
  \op{\N,0} \exp\p{it\Omega_{\N\pm}S_{\N\pm}^\x},
\end{align}
where the Rabi frequencies $\Omega_{\N,+},\Omega_{\N,-}$ and
pseudo-spin operators $S_{\N\pm}^\x$ are respectively given in
\eqref{eq:O_N+}, \eqref{eq:O_N-}, and \eqref{eq:S_NX_x}.  Denoting the
eigenstates of $S_{\N\pm}^\x$ by
\begin{align}
  \ket{\N,\S_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} + \ket{\N\pm}},
  &&
  \ket{\N,\A_\pm} \equiv \f1{\sqrt2} \p{\ket{\N,0} - \ket{\N\pm}},
\end{align}
and defining the identity operator projected to the relevant subspace,
\begin{align}
  \1_{\N\pm} \equiv \op{\N,0} + \op{\N\pm}
  = \op{\N,\S_\pm} + \op{\N,\A_\pm},
\end{align}
we can write the state $\rho_N^{(\pm)}\p{t}$ and excited-state
projectors $\P_{N\pm}$ in the form
\begin{align}
  \rho_N^{(\pm)}\p{t} = \f1{\Tr\P_{N,0}} \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm} + e^{i2t\Omega_{\N\pm}}\op{\N,\S_\pm}{\N,\A_\pm}
    + e^{-i2t\Omega_{\N\pm}}\op{\N,\A_\pm}{\N,\S_\pm}},
  \label{eq:rho_N_pm_t}
\end{align}
and
\begin{align}
  \P_{N\pm} = \sum_{\abs{\N}=N} \f12
  \sp{\1_{\N\pm}-\op{\N,\S_\pm}{\N,\A_\pm}-\op{\N,\A_\pm}{\N,\S_\pm}},
\end{align}
from which it follows that the net excited-state population at time
$t$ is
\begin{align}
  \bk{\P_{N\pm}\p{t}}
  \equiv \Tr\sp{\rho_N^{(\pm)}\p{t}\P_{N\pm}}
  = \f12 - \f12 \bk{\cos\p{2t\Omega_{\N\pm}}}_{\abs{\N}=N},
  \label{eq:P_N_pm_t}
\end{align}
where $\bk{X}_{\abs{\N}=N}\equiv\sum_{\abs{\N}=N}X/\Tr\P_{N,0}$ is an
average of $X$ over all choices of $N$ distinct nuclear spins.
